Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-06-16 Manuscript/Manuscript.Rmd
Ms: correct path for res table

---
title: Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit
author: 
  - name: "Kei Jokura<sup>1,`*`</sup>, Nobuo Ueda<sup>2</sup>, Martin Gühmann<sup>3</sup>, Luis Alfonso Yañez-Guerra<sup>1</sup>, Piotr Słowiński<sup>1</sup>, Kyle C. A. Wedgewood<sup>1</sup> and Gáspár Jékely<sup>1,4,`*`</sup>"
    affiliation: "<sup>1</sup>Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4 4QD, United Kingdom <sup>2</sup>Okinawa Institute of Science and Technology, Okinawa, Japan, <sup>3</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom, <sup>4</sup>Centre for Organismal Studies (COS), University of Heidelberg, 69120 Heidelberg, Germany  <br> <sup>*</sup>Correspondence`:` k.jokura @ exeter.ac.uk, gaspar.jekely @ cos.uni-heidelberg.de"
output:
  html_document:
    theme: lumen
  word_document: default
  pdf_document:
    latex_engine: lualatex
    template: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized from L-arginine by nitric oxide synthase (NOS) is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca^2+^/calmodulin-dependent activation of NOS, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC).
This can occur cell-autonomously or in other cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

@@

Here we uncovered an essential function for NO signalling in larval UV/violet-light avoidance behaviour.
In *Platynereis* larvae, UV/violet avoidance is mediated by brain ciliary photoreceptor cells (cPRCs) and is characterised by downward swimming [@Veraszto2018].
The cPRCs express a ciliary-type opsin, c-opsin1 [@Arendt2004] that forms a UV-absorbing bistable photopigment with an absorption maximum around 384 nm [@tsukamoto2017; @Veraszto2018; @Veedin2021].
Upon UV/violet exposure, the cPRCs show a characteristic biphasic calcium response that is *c-opsin1* dependent.
UV/violet avoidance is also *c-opsin1*-dependent and is defective in *c-opsin1* mutants [@Veraszto2018].
Here we show that *Platynereis* *NOS* is expressed in interneurons of the cPRC circuit and is required for UV/violet-avoidance.
By combining calcium imaging across the fully-mapped cPRC circuit [@Veraszto2018] with genetic perturbations and mathematical modelling, we describe how NO tunes circuit dynamics through non-synaptic retrograde signalling to cPRCs.
This delayed neuroendocrine feedback integrates UV/violet exposure to induce a short-term memory manifested in altered circuit activity and an aversive behavioural response.

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of bilaterian NOS sequences (Figure 1---figure supplement 1).
To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1D and Figure 1---figure supplement 2).
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots (Figure 1---figure supplement 2).
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated-tdTomato (Figure 1E).
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] (Figure 1B, C).
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGWa interneurons, which are also cPRC targets (Figure 1C,F).
The projections of INNOS cells are segregated into input (dendritic) and output (axonal) compartments, with cPRC inputs occurring on the dendritic part in the dense neurosecretory plexus.
INNOS output synapses form in the more ventral projection region (Figure 1---figure supplement 3).
INRGWa neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the cholinergic MC ciliomotor neuron and cells of the prototroch ciliary band (Figure 1C,F) [@veraszto2017].

```{r echo=FALSE,fig.cap="**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(E)** Expression of a *NOS* reporter (NOSp::palmi-3xHA: magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(F)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses."}

@@

We next tested whether NO signalling regulates UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early stop codon and thus likely representing null alleles (Figure 3---figure supplement 1A).
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene in *Platynereis*.
To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light (Figure 3A and Figure 3---figure supplement 1B).
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 3A, B and Figure 3---figure supplement 1B, C).
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae [@Veraszto2018] and reveals a requirement for NOS in UV-avoidance behaviour.
Wild type but not mutant larvae also showed an increase in swimming speed under UV light that may be due to downward swimming trajectories (swimming in the direction of gravity) (Figure 3B and Figure 3---figure supplement 1C).
We also tested directional phototaxis, by exposing larvae to 480 nm directional collimated light from the top of the column.
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] (Figure 3D and Figure 3---figure supplement 1G).

To distinguish between an acute and developmental function of NOS in light responses, we also tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance.
In contrast, phototaxis was not affected (Figure 3C, E).
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system, reminiscent of the function of NO signalling in *Drosophila* eye development [@Gibbs1998a].

@@

To investigate how NO signalling alters the dynamics of the cPRC circuit, we carried out calcium imaging experiments.
We ubiquitously expressed the calcium sensor GCaMP6s in larvae and imaged calcium signals during 405 nm stimulation of the cPRCs.
As we have shown previously, a 20-sec local stimulation of cPRC cilia led to a transient increase in cPRC calcium levels, followed by a transient decrease [@Veraszto2018].
After \~20 sec, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus -- a response that may involve depolarisation (Figure 4A, B).
This activation phase occurs after the 20 sec stimulation period and is likely due to a delayed neuroendocrine feedback [@Veraszto2018].
To determine whether NO mediates such a feedback, we repeated the experiment in *NOS*-mutant larvae.
While we detected the initial activation phase followed by inhibition, in homozygous *NOS*-mutants for both CRISPR alleles this was not followed by delayed activation.
Instead, calcium levels dropped to a low steady-state level (Figure 4A, B).
We thus identified a requirement for NO signalling in the late-phase activation of cPRCs.

@@

We aimed next to identify the NO receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in the cPRCs [@veraszto2017] (not shown).
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1 and 2).
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree (Figure 4---figure supplement 1 and 2).

@@

To investigate how NO and NIT-GC signalling influence the dynamics of the cPRC circuit, we imaged calcium signals from postsynaptic neurons in wild type, mutant and morphant larvae.
We were able to image the activity of all neurons in the cPRC circuit (INNOS, INRGWa, Ser-h1 and MC).
The MC cell was identified based on its position and intrinsic activity [@veraszto2017].
To unambiguously identify all other cells from which we recorded calcium signals, we developed an on-slide immunostaining method (Figure 5---figure supplement 1A).
We used the cell-specific antibody markers against RYamide (INNOS) (Figure 5---figure supplement 1B-D), RGWamide (INRGWa) and serotonin (Ser-h1) [@Conzelmann2011] to immunostain agar-embedded larvae following calcium imaging.
Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision (Figure 5A, B).
Due to the stereotypy of the larvae, we could also identify neurons based on their position and calcium activity in activity-correlation maps (Figure 5C).

We first quantified the responses of the INNOS and INRGWa interneurons during 405 nm stimulation of the cPRCs.
In both wild type and *NOS*-mutant larvae, INNOS cells showed an increase in calcium during stimulation (Figure 5D).
In contrast, the INNOS response was flat or slightly negative in NIT-GC2 morphant larvae (Figure 5E) revealing an essential role for NIT-GC2-mediated cPRC suppression in INNOS activation.
INRGWa cells were initially inhibited during cPRC stimulation, followed by a delayed activation paralleling the second activation phase of cPRCs.
This late INRGWa response was lacking in *NOS*-mutants (Figure 5F).
In NIT-GC2 morphants, INRGWa cells showed a transient increase in calcium that decayed after light off and a delayed activation was not present (Figure 5G).

Next, we imaged calcium signals from Ser-h1 and MC neurons in wild type and *NOS* mutant larvae.
Ser-h1 cells showed an activation profile that correlated with cPRC activity, including a reduction in calcium during stimulation followed by rebound, a response that was defective in *NOS* mutants (Figure 5I).
MC cells showed sustained activation, including a late-phase that was lacking in *NOS* mutants (Figure 5I).
These data suggest that during 405 nm stimulation the Ser-h1 cells are inhibited and MC cells are activated, and this regulation is NO-dependent.
This pattern is expected to inhibit ciliary activity in the prototroch but not in the other ciliary bands (with no Ser-h1 or MC synapses), triggering NO-dependent downward swimming.

```{r echo=FALSE,fig.cap="**Figure 5. NOS- and NIT-GC2-dependent dynamics of the cPRC circuit.** **(A, B)** GCaMP6s imaging from cPRCs and INNOS cells (left panels) followed by on-slide immunostaining for (A) RYamide to label INNOS and (B) RGWamide+serotonin to label INRGWa and Ser-h1 (red). Nuclei are stained with DAPI (cyan). Asterisks indicate cPRC nuclei. Numbers mark the same cells in the GCaMP and immunostaining images matched by position. **(C)** Correlation map of neuronal activity of the cPRCs, INNOS, INRGWa, Ser-h1 and MC neurons. **(D)** GCaMP6s fluorescence in INNOS cells in wild type (WT) and NOSΔ11/Δ11 mutant larvae during 405 nm stimulation of the cPRC cilia. **(E)** GCaMP6s fluorescence in INNOS cells in NIT-GC2 morphant larvae during 405 nm stimulation. **(F)** GCaMP6s fluorescence in INRGWa cells in wild type and NOSΔ23/Δ23 mutant larvae during 405 nm stimulation. **(G)** GCaMP6s fluorescence in INRGWa cells in NIT-GC2 morphant larvae during 405 nm stimulation. **(H, I)** GCaMP6s fluorescence in (H) Ser-h1 cells and (I) the MC cell in wild type and NOSΔ11/Δ11 mutant larvae during 405 nm stimulation."}

@@

To further analyse the dynamics of responses to UV light and formally describe cPRC phototransduction, synaptic connections and NO retrograde signalling, we developed a mixed cellular-circuit-level mathematical model.
We used our Ca^2+^ imagining data of cPRC, INNOS and INRGWa cells collected in wild type, *NOS* knockout, and *NIT-GC2* morphant larvae to formulate assumptions that are the basis of the proposed model.

We model a direct c-opsin1-dependent response to UV leading to the initial rise in Ca^2+^ levels.
The mechanism of this signal is not known but may include Gα and Gβ signalling, CNGAα channel activation or other pathways.
We further assume that in cPRC cells, Ca^2+^ levels increase with cGMP.
We model a decrease in cGMP and Ca^2+^ levels due to a NIT-GC2-dependent pathway (Figre 6A).
The effect of NO in cPRC cells is captured by another term that describes a NIT-GC1 and NO-dependent increase in cGMP.
Based on the synaptic connectome, we infer a feedforward coupling between cPRC cells and INNOS and INRGWa cells.
The proposed synaptic coupling is UV-dependent, decays linearly and is suppressed in NIT-GC2 morphants (Figure 6A).

In the INNOS cells, we assume that excitatory synaptic input from cPRC (or rebound from tonic inhibition) leads to a rise in Ca^2+^ leading to NO production via NOS.
This input is NIT-GC2-dependent.
In the INRGWa cells, we model a decrease in Ca^2+^ based on an inhibitory UV-dependent synaptic signal from the cPRCs.
We further assume the existence of direct excitatory coupling between cPRC Ca^2+^ levels and INRGWa Ca^2+^ to account for the late effects of cPRC on INRGWa (Figure 6A).
The UV-dependent and Ca^2+^-dependent signals may be mediated by different transmitters released by cPRCs during distinct phases of their activation cycle.
In addition, we assume a direct inhibitory coupling between INNOS Ca^2+^ and INRGWa Ca^2+^.
Finally, in all variables, we assumed a linear decay and constant production to set a steady state (Figure 6A).
Since the aim of the model is to capture the normalised fluorescence data, the model is nondimensionalised.
UV stimulation is modelled as a square pulse.

To find model parameters producing an output fitting the experimental data we employed a global optimisation method known as a genetic algorithm (GA).
Through this optimisation procedure, we found parameters that gave a very good fit to our experimental data under all conditions.
We could fit to mean Ca^2+^ values but also individual calcium transients (Figure 6B, Figure 6--figure supplement 1).
Our model only includes a minimal set of parameters and assumptions of interactions that are required to describe the dynamics of cPRC, INNOS and INRGWa in wild type and loss-of-function conditions (Figure 6A, B).
The model highlights that cPRC phototransduction employs distinct pathways that operate on different time scales and differentially influence cPRC Ca^2+^ levels.
The coupling between cPRCs and interneurons also requires different signalling mechanisms, either through distinct neurotransmitters or receptors expressed in the different cells.

>>>>

---
title: Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit
author: 
  - name: "Kei Jokura<sup>1,`*`</sup>, Nobuo Ueda<sup>2</sup>, Martin Gühmann<sup>3</sup>, Luis Alfonso Yañez-Guerra<sup>1</sup>, Piotr Słowiński<sup>1</sup>, Kyle C. A. Wedgewood<sup>1</sup> and Gáspár Jékely<sup>1,4,`*`</sup>"
    affiliation: "<sup>1</sup>Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4 4QD, United Kingdom <sup>2</sup>Okinawa Institute of Science and Technology, Okinawa, Japan, <sup>3</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom, <sup>4</sup>Centre for Organismal Studies (COS), University of Heidelberg, 69120 Heidelberg, Germany  <br> <sup>*</sup>Correspondence`:` k.jokura @ exeter.ac.uk, gaspar.jekely @ cos.uni-heidelberg.de"
output:
  html_document:
    theme: lumen
  pdf_document:
    latex_engine: lualatex
    template: default
  word_document: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized from L-arginine by nitric oxide synthase (NOS) is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca2+/calmodulin-dependent activation of NOS, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC).
This can occur cell-autonomously or in other cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

@@

Here we uncovered an essential function for NO signalling in larval UV/violet-light avoidance behaviour.
In *Platynereis* larvae, UV/violet avoidance is mediated by brain ciliary photoreceptor cells (cPRCs) and is characterised by downward swimming [@Veraszto2018].
The cPRCs express a ciliary-type opsin, c-opsin1 [@Arendt2004] that forms a UV-absorbing bistable photopigment with an absorption maximum around 384 nm [@tsukamoto2017; @Veraszto2018; @Veedin2021].
Upon UV/violet exposure, the cPRC show a characteristic biphasic calcium response that is *c-opsin1* dependent.
UV/violet avoidance is also *c-opsin1*-dependent and is defective in *c-opsin1* mutants [@Veraszto2018].
Here we show that *Platynereis* *NOS* is expressed in interneurons of the cPRC circuit and is required for UV/violet-avoidance.
By combining calcium imaging across the fully-mapped cPRC circuit [@Veraszto2018] with genetic perturbations and mathematical modelling, we describe how NO tunes circuit dynamics through non-synaptic retrograde signalling to cPRCs.
This delayed neuroendocrine feedback integrates UV/violet exposure to induce a short-term memory manifested in altered circuit activity and an aversive behavioural response.

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of bilaterian NOS sequences (Figure 1---figure supplement 1).
To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1D and Figure 1---figure supplement 2).
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots (Figure 1---figure supplement 2).
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated-tdTomato (Figure 1E).
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] (Figure 1B,C).
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Figure 1C,F).
The projections of INNOS cells are segregated into input (dendritic) and output (axonal) compartments, with cPRC inputs occurring on the dendritic part in the dense neurosecretory plexus.
INNOS output synapses form in the more ventral projection region (Figure 1---figure supplement 3).
INRGW neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the prototroch ciliary band and the cholinergic MC ciliomotor neurons (Figure 1C,F) [@veraszto2017].

```{r echo=FALSE,fig.cap="**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGW, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(E)** Expression of a *NOS* reporter (NOSp::palmi-3xHA: magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(F)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses."}

@@

We next tested whether NO signalling regulates UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early stop codon and thus likely representing null alleles (Figure 3---figure supplement 1A).
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene in *Platynereis*.
To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light (Figure 3A and Figure 3---figure supplement 1B).
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 3A, B and Figure 3---figure supplement 1B,C).
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae [@Veraszto2018] and reveals a requirement for NOS in UV-avoidance behaviour.
Wild type but not mutant larvae also showed an increase in swimming speed under UV light that may be due to downward swimming trajectories (swimming in the direction of gravity) (Figure 3B and Figure 3---figure supplement 1C).
We also tested directional phototaxis, by exposing larvae to 480 nm directional collimated light from the top of the column.
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] (Figure 3D and Figure 3---figure supplement 1G).

To distinguish between an acute and developmental function of NOS in light responses, we next tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance.
In contrast, phototaxis was not affected (Figure 3C, E).
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system, reminiscent of the function of NO signalling in *Drosophila* eye development [@Gibbs1998a].

@@

To investigate how NO signalling alters the dynamics of the cPRC circuit, we carried out calcium imaging experiments.
We ubiquitously expressed the calcium sensor GCaMP6s in larvae and imaged calcium signals during 405 nm stimulation of the cPRCs.
As we have shown previously, a 20-sec local stimulation of cPRC cilia led to a transient increase in cPRC calcium levels, followed by a transient decrease [@Veraszto2018].
After \~20-sec, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus -- a response that may involve depolarisation (Figure 4A, B).
This activation phase occurs after the 20 sec stimulation period and is likely due to a delayed neuroendocrine feedback [@Veraszto2018].
To determine whether NO mediates such a feedback, we repeated the experiment in *NOS*-mutant larvae.
While we detected the initial activation phase followed by inhibition, in homozygous *NOS*-mutants for both CRISPR alleles this was not followed by delayed activation.
Instead, calcium levels dropped to a low steady-state level (Figure 4A, B).
We thus identified a requirement for NO signalling in the late-phase activation of cPRCs.

@@

We aimed next to identify the NO receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@veraszto2017].
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1 and 2).
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree (Figure 4---figure supplement 1 and 2).

@@

To investigate how NO and NIT-GC signalling influence the dynamics of the cPRC circuit, we imaged calcium signals from postsynaptic neurons in wild type, mutant and morphant larvae.
We were able to image the activity of all neurons in the cPRC circuit (INNOS, INRGW, Ser-h1 and MC).
The MC cell was identified based on its position and intrinsic activity [@veraszto2017].
To unambiguously identify all other cells from which we recorded calcium signals, we developed an on-slide immunostaining method (Figure 5---figure supplement 1A).
We used the cell-specific antibody markers against RYamide (INNOS) (Figure 5---figure supplement 1B-D), RGWamide (INRGW) and serotonin (Ser-h1) [@Conzelmann2011] to immunostain agar-embedded larvae following calcium imaging.
Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision (Figure 5A,B).
Due to the stereotypy of the larvae, we could also identify neurons based on their position and calcium activity in activity-correlation maps (Figure 5C).

We first quantified the responses of the INNOS and INRGW interneurons during 405 nm stimulation of the cPRCs.
In both wild type and *NOS*-mutant larvae, INNOS cells showed an increase in calcium during stimulation (Figure 5D).
In contrast, the INNOS response was flat or slightly negative in NIT-GC2 morphant larvae (Figure 5E) revealing an essential role for NIT-GC2-mediated cPRC suppression in INNOS activation.
INRGW cells were initially inhibited during cPRC stimulation, followed by a delayed activation paralleling the second activation phase of cPRCs.
This late INRGW response was lacking in *NOS*-mutants (Figure 5F).
In NIT-GC2 morphants, INRGW cells showed a transient increase in calcium that decayed after light off and a delayed activation was not present (Figure 5G).

Next, we imaged calcium signals from Ser-h1 and MC neurons in wild type and *NOS* mutant larvae.
Ser-h1 cells showed an activation profile that correlated with cPRC activity, including a reduction in calcium during stimulation followed by rebound, a response that was defective in *NOS* mutants (Figure 5I).
MC cells showed sustained activation, including a late-phase that was lacking in *NOS* mutants (Figure 5I).
These data suggest that during 405 nm stimulation the Ser-h1 cells are inhibited and MC cells are activated, and this regulation is NO-dependent.
This pattern is expected to inhibit ciliary activity in the prototroch but not in the other ciliary bands, triggering NO-dependent downward swimming.

```{r echo=FALSE,fig.cap="**Figure 5. NOS- and NIT-GC2-dependent dynamics of the cPRC circuit.** **(A, B)** GCaMP6s imaging from cPRCs and INNOS cells (left panels) followed by on-slide immunostaining for (A) RYamide to label INNOS and (B) RGWamide+serotonin to label INRGW and Ser-h1 (red). Nuclei are stained with DAPI (cyan). Asterisks indicate cPRC nuclei. Numbers mark the same cells in the GCaMP and immunostaining images matched by position. **(C)** Correlation map of neuronal activity of the cPRCs, INNOS, INRGW, Ser-h1 and MC neurons. **(D)** GCaMP6s fluorescence in INNOS cells in wild type (WT) and NOSΔ11/Δ11 mutant larvae during 405 nm stimulation of the cPRC cilia. **(E)** GCaMP6s fluorescence in INNOS cells in NIT-GC2 morphant larvae during 405 nm stimulation. **(F)** GCaMP6s fluorescence in INRGW cells in wild type and NOSΔ23/Δ23 mutant larvae during 405 nm stimulation. **(G)** GCaMP6s fluorescence in INRGW cells in NIT-GC2 morphant larvae during 405 nm stimulation. **(H, I)** GCaMP6s fluorescence in (H) Ser-h1 cells and (I) the MC cell in wild type and NOSΔ11/Δ11 mutant larvae during 405 nm stimulation."}

@@

To further analyse the dynamics of responses to UV light and formally describe cPRC phototransduction, synaptic connections and NO retrograde signalling, we developed a mixed cellular-circuit-level mathematical model.
We used our Ca2+ imagining data of cPRC, INNOS and INRGW cells collected in wild type, *NOS* knockout, and *NIT-GC2* morphant larvae to formulate assumptions that are the basis of the proposed model.

We model a direct c-opsin1-dependent response to UV leading to the initial rise in Ca2+ levels.
The mechanism of this signal is not known but may include Gα and Gβ signalling, CNGAα channel activation or other pathways.
We further assume that in cPRC cells, Ca2+ levels increase with cGMP.
We model a decrease in cGMP and Ca2+ levels due to a NIT-GC2-dependent pathway (Figure 6A).
The effect of NO in cPRC cells is captured by another term that describes a NIT-GC1 and NO-dependent increase in cGMP.
Based on the synaptic connectome, we infer a feedforward coupling between cPRC cells and INNOS and INRGW cells.
The proposed synaptic coupling is UV-dependent, decays linearly and is suppressed in NIT-GC2 morphants (Figure 6A).

In the INNOS cells, we assume that excitatory synaptic input from cPRC (or rebound from tonic inhibition) leads to a rise in Ca2+ leading to NO production via NOS.
This input is NIT-GC2-dependent.
In the INRGW cells, we model a decrease in Ca2+ based on an inhibitory UV-dependent synaptic signal from the cPRCs.
We further assume the existence of direct excitatory coupling between cPRC Ca2+ levels and INRGW Ca2+ to account for the late effects of cPRC on INRGW.
The UV-dependent and Ca2+-dependent signals may be mediated by different transmitters released by cPRCs during distinct phases of their activation cycle.
In addition, we assume a direct inhibitory coupling between INNOS Ca2+ and INRGW Ca2+.
Finally, in all variables, we assumed a linear decay and constant production to set a steady state (Figure 6A).
Since the aim of the model is to capture the normalised fluorescence data, the model is nondimensionalised.
UV stimulation is modelled as a square pulse.
To find model parameters producing an output fitting the experimental data we employed a global optimisation method known as a genetic algorithm (GA).

Our model only includes a minimal set of parameters and assumptions of interactions that are required to describe the dynamics of cPRC, INNOS and INRGW in wild type and loss-of-function conditions (Figure 6A, B).
The model highlights that cPRC phototransduction employs distinct pathways that operate on different time scales and differentially influence cPRC Ca2+ levels.
The coupling between cPRCs and interneurons also requires different signalling mechanisms, either through distinct neurotransmitters or receptors expressed in the different cells.

To identify possible molecular pathways, we analysed single-cell transcriptome data for cPRC, INNOS and INRGW identified by spatial mapping [@Achim2015] and unique marker genes [@williams2017].
In cPRCs, we found transporters and synthetic enzymes indicating cholinergic, glutaminergic, glycinergic, GABAergic and adrenergic neurotransmission (Figure 6C).
For each neurotransmitter, we also found receptor-encoding genes expressed in INNOS and INRGW (Figure 6C).
The two types of interneurons often expressed different subunits or types of these receptors, indicating differential signalling (Figure 6C).
